Improving the Stability of Type Soundness Proofs in Dafny

### 2023-10-17 paper/paper.tex
what

\section{Introduction}
Type soundness proofs for strongly-typed programming languages are often carried
out in type-theory based proof assistants like Coq or Agda. However, recent
small-scale experiements \cite{mayer-blog-post} and large-scale mechanized
metatheory developments \cite{cedar-proofs}.  have shown that Dafny is capable
enough as a proof assistant to encode type soundness proofs. In these
developments, Dafny serves both as the host language in which the target
language can be implemented, and a program verifier in which the mechanized
proof can be completed.

The usual way that a target language semantics is implemented in Dafny
is by way of a definitional interpreter \cite{definitional-interpreters}, a function \texttt{eval} which takes an
an abstract syntax tree \texttt{e} which represents a term in the target language, and returns either an error,
or the value \texttt{v} the term evaluates to. In principle, this implements some abstract semantics defined
by inference rules on paper: a semantic judgment $\texttt{e} \downarrow \texttt{v}$ holds if and only if \texttt{eval(e) = Ok(v)}.

@@

The goal of a type soundness proof is then to demonstrate that well-typed
programs do not go wrong: if $\vdash \texttt{e} : \texttt{t}$, then $\texttt{e}
\downarrow \texttt{v}$ for some value \texttt{v} of type \texttt{t}. When proven
on paper, this proof proceeds by an induction on the typing derivation of
$\vdash \texttt{e} : \texttt{t}$. For each rule defining the typing judgment, we
prove that if the premises do not go wrong, then the conclusion also does not go
wrong. Meanwhile, a \emph{mechanized} proof of this fact in Dafny takes the form
of a lemma which requires \texttt{check(e,t)}, and ensures that \texttt{eval(e)
= Ok(v)} and that \texttt{v} has type \texttt{t}. The proof of this lemma
necessarily looks like a pattern-match on the term \texttt{e}, making recursive
calls on the subterms to establish the inductive hyoptheses. For syntax-directed type systems
of the sort we will consider in this extended abstract\footnote{Strictly speaking, this is not a restriction of our tehcnique. The only practical way to implement a non-syntax-directed type system is to build a syntax-directed version, prove the two versions equivalent, and then implement the syntax-directed one.},
the structure of these two kinds of proofs are the same: each case of the pattern match encodes the proof for the corresponding typing rule for the corresponding syntactic form.

As such, the size of a type soundness proof in Dafny grows linearly with the
size of the language; more language features and more expressions mean more
cases.  This can make it hard to ensure that the proof is \emph{stable}.  Unlike
mechanized metatheory developments in tools backed by a type-theoretic proof
checker, Dafny's SMT-backed verification can be \emph{unstable}: the result of
verifying the type soundness theorem can change, seemingly at random.
Unfortunately, followign the proof structure described above leads to very unstable proofs in Dafny. Anecdotally, we have
found it impossible to scale this kind of proof to anything approximating a realistic language without encountering enormous
proof instability barriers.

In this extended abstract, we propose a new proof structure for mechanized type soundness proofs in Dafny which ensures proof stability.
By giving partial specifications of the evaluator and type checker, and proving type soundness relative to those specifications, we
can eliminate much of the \jwc{words}
Our technique has been used to verify \jwc{word}

In Section~\ref{sec:sts}, we present the technique by walking through a traditional-style proof of type soundness in Dafny.
Then in Section~\ref{sec:eval}, we evaluate the technique comparing empirical metrics of proof stability against other methods of structuring type soundness proofs, and other ways of writing a typechecker.
Lastly, in Section~\ref{sec:discussion}, we discuss how our technique has been deployed in practice, speculate about how it can be applied to even more complex languages,
and discuss the limitations and drawbacks.

\section{Stable Type Soundness Proofs}
\label{sec:sts}
For a running example, let's consider the first-order expression language we
use for our evaluation in Section~\ref{sec:eval}.  The language has
types for integers, booleans, and records, all given as the Dafny type
\texttt{Ty}.  For terms, we have variables (represented as strings, since the
langauge has no binders), addition, subtraction, division, and, or,
conditionals, and record expressions and projections. These are all represented
by a Dafny ADT \texttt{Term}, which has a variant for each syntactic form.
To demonstrate how the language works, and how our technique applies, we focus
on just the behavior of the addition operation.

@@

\subsection*{Finding Stability}
Unfortunately, as we'll see in Section~\ref{sec:eval}, this proof approach is
not stable. We realized this while attemping to scale this style
of proof up to a realistic language. With enough operations and language
features, the proof resource usage varies wildly between runs,
enough that the verification will fail at random. At its core, this
instability arises from all of the afformentioned unguided reasoning the solver has to do about the
code of the typechecker and evaluator,
to get from the premise of the lemma to the premises of the IH, and then
from the conclusions of the IH to the conclusion of the case.
A patchwork solution is to add guidance in the form of assertions around the
recursive calls, to spell out the intermediate steps more directly.
\jwc{Words, about why this doesn't work in the long run}

A common and more robust solution to dealing with proof instability is to specify
the functions involved in the proof, and then make the functions themselves opaque.
This way, the solver can only interact with the functions though the specification,
cutting down the search space of possible proofs and hence improving stability.
Moreover, if the specifications are written in the form of separate lemmas,
the programmer can control how and when different parts of the function's sepcification
is revealed to the solver.

But which functions to make opaque, and how to specify them?
Inspiration comes from noticing that \texttt{sound} depends only on
facts about the safety predicate \texttt{isSafe}, and not directly
on facts about evaluation. In fact, the only facts it needs about \texttt{isSafe}
are that in every case, the results of the IH calls jointly imply the conclusion.
To illustrate, the addition case for type soundness holds
for \emph{any} safety predicate \texttt{isSafe} --- even an opaque one --- which has the property
that \texttt{isSafe(env,e1,IntTy)} and \texttt{isSafe(env,e2,IntTy)}
together imply \texttt{isSafe(env,Add(e1,e2),IntTy)}.

@@

\begin{itemize}
  \item (\texttt{lang.dfy}): The langauge and type definitions, common to all typecheckers and the evaluator.
  \item (\texttt{eval.dfy}): Contains the evaluator for our expressiong language. All of the type soundness proofs are relative to this evaluator.
  \item (\texttt{monadic.dfy}): This is the typechecker and type safety proof described in
  the first part of Section~\ref{sec:sts}.  The typechecker uses one main method
  \texttt{infer} and some helper functions to infer a type for a term, and then checks
  that it's equal to the given type. \jwc{Go back and make this true, in the code}
  \item (\texttt{direct}): This is the same as the typechecker in \texttt{monadic.dfy}, with all of the
  syntactic sugar for \texttt{Result}-binds (\texttt{:-}) unfolded, and all of the helper
  functions inlined. We include this file to measure the degree to which adding
  layers of abstraction hurts proof stability. The proof is essentially the same as the one in \texttt{monadic.dfy}.
  \item (\texttt{bidirectional.dfy}): This typechecker is written in the same style as the typechecker in \texttt{monadic.dfy}. The typing relation that it impelents is slightly different,
        however, including a subtyping rule which implements record width and depth subtyping.
        The type soundness proof changes to include calls to reflexivity and transitivity lemmas about subtyping, which cannot be inferred by the solver.
  \item (\texttt{semantic.dfy} \jwc{fix}):
        This file implements the first half of our technique, making the safety
        property opaque, proving compatability lemmas about the evaluator, and
        calling them in the soundness proof.  The typechecker is identical to
        the one in \texttt{bidirectional.dfy}.
  \item (\texttt{semantic-invert.dfy}): This file implements the full stabilizing technique, adding
  \item (\texttt{std.dfy} and \texttt{util.dfy}): Auxiliary functions and definitions not specific to the language's evaluator or typechecker.
\end{itemize}

We run the experiment by running the built-in \texttt{dafny measure-complexity}
command on the files containing each type soundness proof, with the flags \texttt{--iterations:250 --log-format csv}.
Each invocation of \texttt{measure-complexity} command verifies the file $250$ times, and dumps
the verification results, durations, and resource usages to a CSV file. We then parse and analyze the CSV
file with a python script, which computes the means and standard deviations of verification duration and resource usage for the solver's task of proving correctness
of the VCs in the file's main soundness theorem.
The experiments were run on a 2023 MacBook Pro with an Apple Silicon M2 processor, and 32GB memory, runnning Dafny v4.3.0, and z3 v4.12.1.

@@

Semantically though, Cedar is not \emph{all} that much more complex to model than the
toy language we evaluate for this extended abstract. Like our toy language,
Cedar's evaluator is simplified by the fact that it is terminating, first order, and has no binders.  Although we have not tested it, our
technique should scale to languages with nontermination --- by step-indexing the
evaluator and proving type-safety with failure to converge in $n$ steps being
considered ``safe'' \jwc{cite step-indexing} --- and higher-order functions ---
by applying standard tricks to handle variable binding. \jwc{cite}

\subsubsection*{Manual Effort \& Benefit}
All of this benefit does come at the cost of some of the automation that Dafny
users expect. In taking possible work away from the solver to make the proof
more stable, we are necessarily creating more work for ourselves!  As we saw in
the previous section, the type soundness proofs in the last two cases spell out
many more steps along the way.

>>>>

\section{Introduction}
Type soundness proofs for strongly-typed programming languages are often carried
out in type-theory based proof assistants like Coq or Agda. However, recent
small-scale experiements \cite{mayer-blog-post} and large-scale mechanized
metatheory developments \cite{cedar-dafny}.  have shown that Dafny is capable
enough as a proof assistant to encode type soundness proofs. In these
developments, Dafny serves both as the host language in which the target
language can be implemented, and a program verifier in which the mechanized
proof can be completed.

\jwc{Put the math on a different line: let it breathe. De-convolve the sentences.}
The usual way that a target language semantics is implemented in Dafny
is by way of a definitional interpreter \cite{definitional-interpreters}, a function \texttt{eval} which takes an
an abstract syntax tree \texttt{e} which represents a term in the target language, and returns either an error,
or the value \texttt{v} the term evaluates to. In principle, this implements some abstract semantics defined
by inference rules on paper: a semantic judgment $\texttt{e} \downarrow \texttt{v}$ holds if and only if \texttt{eval(e) = Ok(v)}.

@@

The goal of a type soundness proof is then to demonstrate that well-typed
programs do not go wrong:
$$\vdash \texttt{e} : \texttt{t} \implies \exists \texttt{v}. \left(\texttt{e} \downarrow \texttt{v} \wedge \texttt{v} : \texttt{t}\right)$$
for some value \texttt{v} of type \texttt{t}. When proven
on paper, this proof proceeds by an induction on the typing derivation of
$\vdash \texttt{e} : \texttt{t}$. For each rule defining the typing judgment, we
prove that if the premises do not go wrong, then the conclusion also does not go
wrong. Meanwhile, a \emph{mechanized} proof of this fact in Dafny takes the form
of a lemma which requires \texttt{check(e,t)}, and ensures that \texttt{eval(e)
= Ok(v)} and that \texttt{v} has type \texttt{t}. The proof of this lemma
necessarily looks like a pattern-match on the term \texttt{e}, making recursive
calls on the subterms to establish the inductive hyoptheses. For syntax-directed type systems
of the sort we will consider in this extended abstract\footnote{Strictly speaking, this is not a restriction of our technique. By far the most common way to implement a non-syntax-directed type system is to build a syntax-directed version, prove the two versions equivalent, and then implement the syntax-directed one.},
the structure of these two kinds of proofs are the same: each case of the pattern match encodes the proof for the corresponding typing rule for the corresponding syntactic form.

The size of a type soundness proof in Dafny grows linearly with the
size of the language; more language features and more expressions mean more
cases.  This can make it hard to ensure that the proof is \emph{stable}.  Unlike
mechanized metatheory developments in tools backed by a type-theoretic proof
checker, Dafny's SMT-backed verification can be \emph{unstable}: the result of
verifying the type soundness theorem can change, from verified to unverified, seemingly at random. 
% \jwc{Jump up and down more about this! The solver may fail, sporadically, without warning. this is catastrophic}
Unfortunately, folowing the proof structure described above leads to very unstable proofs in Dafny. Anecdotally, we have
found it impossible to scale this kind of proof to anything approximating a realistic language without encountering enormous
proof instability barriers.

In this extended abstract, we give a recipe for mechanized type soundness proofs
in Dafny which ensures proof stability.  By giving partial specifications of the
evaluator and type checker, and proving type soundness relative to those
specifications, we can eliminate much of the variance in resource usage.  Our
technique also scales to realistic-size languages, as witnessed by its use to
verify the type soundness proof of the Cedar Language \cite{cedar}.

In Section~\ref{sec:sts}, we present the technique by walking through a traditional-style proof of type soundness in Dafny.
Then in Section~\ref{sec:eval}, we evaluate the technique comparing empirical metrics of proof stability against other methods of structuring type soundness proofs, and other ways of writing a typechecker.
Lastly, in Section~\ref{sec:discussion}, we discuss how our technique has been deployed in practice, speculate about how it can be applied to even more complex languages including Cedar,
and discuss the limitations and drawbacks.

\section{Stable Type Soundness Proofs}
\label{sec:sts}
For a running example, let's consider the first-order expression language we
use for our evaluation in Section~\ref{sec:eval}. The full language definition, as well as the
interpreter and all of the different typecheckers we evaluate can be found in the attached supplementary materials. The language has
types for integers, booleans, and records, all given as the Dafny type
\texttt{Ty}.  For terms, we have variables (represented as strings, since the
langauge has no binders), addition, subtraction, division, and, or,
conditionals, and record expressions and projections. These are all represented
by a Dafny ADT \texttt{Term}, which has a variant for each syntactic form.
To demonstrate how the language works, and how our technique applies, we focus
on just the behavior of the addition operation.

@@

\subsection*{Finding Stability}
Unfortunately, as we'll see in Section~\ref{sec:eval}, this proof approach is
not stable. We realized this while attemping to scale this style
of proof up to a realistic language. With enough operations and language
features, the proof resource usage varies wildly between runs,
enough that the verification will fail at random. At its core, this
instability arises from all of the afformentioned unguided reasoning the solver has to do about the
code of the typechecker and evaluator,
to get from the premise of the lemma to the premises of the IH, and then
from the conclusions of the IH to the conclusion of the case.
A patchwork solution is to add guidance in the form of assertions around the
recursive calls, to spell out the intermediate steps more directly.

The common folklore solution to dealing with proof instability is to specify the
functions involved in the proof, and then make the functions themselves opaque.
This way, the solver can only interact with the functions though the
specification, cutting down the search space of possible proofs and hence
improving stability.  Moreover, if the specifications are written in the form of
separate lemmas, the programmer can control how and when different parts of the
function's sepcification is revealed to the solver. In this case, the dilemma is
that it's not at all clear which functions to make opaque, and how we should
specify them.

Inspiration comes from noticing that \texttt{sound} depends only on
facts about the safety predicate \texttt{isSafe}, and not directly
on facts about evaluation. In fact, the only facts it needs about \texttt{isSafe}
are that in every case, the results of the IH calls jointly imply the conclusion.
To illustrate, the addition case for type soundness holds
for \emph{any} safety predicate \texttt{isSafe} --- even an opaque one --- which has the property
that \texttt{isSafe(env,e1,IntTy)} and \texttt{isSafe(env,e2,IntTy)}
together imply \texttt{isSafe(env,Add(e1,e2),IntTy)}.

@@

\begin{itemize}
  \item (\texttt{lang.dfy}): The langauge and type definitions, common to all typecheckers and the evaluator.
  \item (\texttt{eval.dfy}): Contains the evaluator for our expressiong language. All of the type soundness proofs are relative to this evaluator.
  \item (\texttt{monadic.dfy}): This is the typechecker and type safety proof described in
  the first part of Section~\ref{sec:sts}.  The typechecker uses one main method
  \texttt{infer} and some helper functions to infer a type for a term, and then checks
  that it's equal to the given type.
  \item (\texttt{direct}): This is the same as the typechecker in \texttt{monadic.dfy}, with all of the
  syntactic sugar for \texttt{Result}-binds (\texttt{:-}) unfolded, and all of the helper
  functions inlined. We include this file to measure the degree to which adding
  layers of abstraction hurts proof stability. The proof is essentially the same as the one in \texttt{monadic.dfy}.
  \item (\texttt{bidirectional.dfy}): This typechecker is written in the same style as the typechecker in \texttt{monadic.dfy}. The typing relation that it impelents is slightly different,
        however, including a subtyping rule which implements record width and depth subtyping.
        The type soundness proof changes to include calls to reflexivity and transitivity lemmas about subtyping, which cannot be inferred by the solver.
  \item (\texttt{semantic.dfy}):
        This file implements the first half of our technique, making the safety
        property opaque, proving compatability lemmas about the evaluator, and
        calling them in the soundness proof.  The typechecker is identical to
        the one in \texttt{bidirectional.dfy}.
  \item (\texttt{semantic-invert.dfy}): This file implements the full stabilizing technique, adding
  \item (\texttt{std.dfy} and \texttt{util.dfy}): Auxiliary functions and definitions not specific to the language's evaluator or typechecker.
\end{itemize}

We run the experiment by running the built-in \texttt{dafny measure-complexity}
command on the files containing each type soundness proof, with the flags \texttt{--iterations:250} and \texttt{--log-format csv}.
Each invocation of \texttt{measure-complexity} command verifies the file $250$ times, and dumps
the verification results, durations, and resource usages to a CSV file. We then parse and analyze the CSV
file with a python script, which computes the means and standard deviations of verification duration and resource usage for the solver's task of proving correctness
of the VCs in the file's main soundness theorem.
The experiments were run on a 2023 MacBook Pro with an Apple Silicon M2 processor, and 32GB memory, runnning Dafny v4.3.0, and z3 v4.12.1.

@@

Semantically though, Cedar is not \emph{all} that much more complex to model than the
toy language we evaluate for this extended abstract. Like our toy language,
Cedar's evaluator is simplified by the fact that it is terminating, first order, and has no binders.  Although we have not tested it, our
technique should scale to languages with nontermination --- by step-indexing the
evaluator and proving type-safety with failure to converge in $n$ steps being
considered ``safe'' \cite{step-indexing} --- and higher-order functions ---
by applying standard tricks to handle variable binding.

\subsubsection*{Manual Effort \& Benefit}
All of this benefit does come at the cost of some of the automation that Dafny
users expect. In taking possible work away from the solver to make the proof
more stable, we are necessarily creating more work for ourselves!  As we saw in
the previous section, the type soundness proofs in the last two cases spell out
many more steps along the way. Anecdotally, however, we have found it practically impossible
to scale a proof that doesn't use this technique. So while this technique does require more manual effort,
we have found it to be the only way to structure a type soundness proof that does not encounter impassable instability obstacles.
